\documentclass[12pt,a4paper]{report}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{csquotes}
\usepackage[backend=biber,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{setspace}
\usepackage{parskip}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  pdftitle={Toward a Probabilistic Semantics for the Lambek Calculus},
  pdfauthor={}
}

\onehalfspacing

\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\newcommand{\LC}{\mathsf{L}}
\newcommand{\At}{\mathsf{At}}
\newcommand{\Frag}{\mathsf{Frag}}
\newcommand{\Lex}{\mathsf{Lex}}
\newcommand{\Sem}{\models_{\mathcal C}}
\newcommand{\concat}{\mathbin{\cdot}}
\newcommand{\Pow}{\mathcal{P}}

\title{Probabilistic Semantics for Lambek Calculus\\\large Supervised by Mehrnoosh Sadrzadeh}
\author{}
\date{April 2026}

\begin{document}

\hypersetup{pageanchor=false}
\pagenumbering{roman}
\maketitle
\setcounter{page}{2}

\begin{abstract}
This thesis studies a corpus-relative probabilistic semantics for the associative Lambek
calculus. The Lambek calculus was introduced by Lambek as a logic of syntactic types and has
since become one of the foundational systems of categorial and type-logical grammar.
Its appeal lies in the tight match between syntax and composition: grammatical derivations
track how word meanings combine, which makes the calculus attractive both in formal
linguistics and in mathematically oriented approaches to natural language processing.

The central motivation of this project is ambiguity. In a purely logical parser, a sentence may
admit several derivations, all equally well-formed at the level of derivability. In practical
language processing, however, one usually wants not merely to decide grammaticality but also
to rank candidate analyses. Probabilistic grammars such as PCFGs and probabilistic CCGs do
exactly this, and their empirical success suggests that a probabilistic version of Lambek-style
parsing is worth developing. The aim here is therefore to define a semantics that assigns to a
string and a Lambek type a value in $[0,1]$, interpreted as the probability or confidence that
the string realises that type relative to a finite corpus.

The thesis contributes a concrete formalisation of this idea. The semantics is restricted to a
finite set of observed corpus fragments, which keeps the recursive clauses well-defined and makes
explicit the sense in which the model is corpus-relative. Particular attention is paid to technical
subtleties in the semantics: words outside the corpus vocabulary are excluded from the formal
domain instead of being given ad hoc values, lexical types not attested for a word receive
probability $0$, and the residual clauses are defined so as to avoid division by zero. The
formalisation also separates direct lexical assignment from the recursive
semantic valuation. On the basis of these definitions, a soundness theorem is proved for the
algebraic axioms of the associative Lambek calculus.

The thesis does \emph{not} claim completeness for the probabilistic semantics introduced here.
On the contrary, it argues that completeness is the natural next problem. Classical completeness
results for the Lambek calculus are already known for several non-probabilistic semantics, but
it remains open whether an analogous theorem can be obtained for the present corpus-based
probabilistic setting. The final chapter therefore outlines future work on completeness,
implementation, smoothing for unseen data, and empirical evaluation against existing
probabilistic parsing frameworks.
\end{abstract}

\tableofcontents
\clearpage
\hypersetup{pageanchor=true}
\pagenumbering{arabic}

\input{chapters/01-introduction}
\input{chapters/02-literature-review}
\input{chapters/03-formalisation}
\input{chapters/04-worked-example}
\input{chapters/05-soundness}
\input{chapters/06-future-work}

\printbibliography

\end{document}
